Equalizer Programming Guide

### 2008-11-10 Developer/ProgrammingGuide/paper.tex
Programming Guide: Threading model updates
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp}

@@

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.15 for Equalizer 0.5.5+svn\\[\medskipamount]
  \today
}

@@

\thispagestyle{empty}
\maketitle

@@

\subsection*{Front Page}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
    1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
    1.3.12  & Oct 16, 2008 & Added \sref{sView}\\
    1.3.13  & Oct 17, 2008 & Minor update in \sref{sMessagePump}\\
    1.3.14  & Oct 21, 2008 & Improved \sref{sConnectionDescription}\\
    1.3.15  & Oct 27, 2008 & Added \sref{sDPlex}\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchonization or protection of shared data.

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

Rendering threads on a single node are by default synchronized. When a
frame is finished, all local rendering threads are done
drawing. Therefore the application can safely modify the data between
the end of a frame and the beginning of a new frame. Furthermore, only
one instance of the application data has to be maintained within a
process since all rendering threads are guaranteed to draw the same frame.

This per-node frame synchronization does not inhibit latency across
rendering nodes. Furthermore, advanced rendering software which
multi-buffers the dynamic parts of the database can disable the per-node
frame synchronization, as explained in \sref{sNodeFrame}. Some scene
graphs for example do implement multi-buffered data, and can profit from
relaxing the frame synchronization.

@@

By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sNodeFrame}.

@@

Equalizer applications use multiple, asynchronous execution threads. The
default execution model is focused on making the porting of existing
applications as easy as possible, as described in \sref{sExecution}. The
default, per-node thread synchronization provided by Equalizer can
gradually be relaxed by advanced applications to gain better performance
through higher asynchronicity.

This section explains the multi-threading in Equalizer in detail and
gives advice on how to optimize applications for performance.

@@

\subsubsection{\label{sNodeFrame}Thread Synchronization Model}

The node main thread and the pipe threads are synchronized with each
other, so that all draw task methods of one rendered frame are executed
at the same time. This model allows to use the same database for
rendering, and safe modifications of this database are possible from the
node thread, since the pipe threads do not execute any rendering tasks
between frames.

Please note that the default thread synchronization synchronizes all
\textsf{Channel::frame\-Draw} operations on a \textbf{single} node with
the node's main thread. The per-node frame synchronization does not
break the asynchronous execution across nodes. Advanced applications can
remove the per-node frame synchronization.

@@

The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model, Equalizer uses the task
method \textsf{frameDrawFinish} which is called on each resource
after the last \textsf{Channel::frameDraw} invocation for this
frame. Consequently, \textsf{Pipe::frameDrawFinish} calls
\textsf{Pipe::releaseFrame\-Lo\-cal} to signal that it is done drawing the
current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

\fig{fFrameSync} illustrates the local frame synchronization. By
removing the calls to \textsf{waitFrameLocal}, the node thread can
disable the synchronization with the pipe threads, which causes the node
operations to overlap with the rendering.

@@

\fig{fMainLoop} outlines the synchronization for the application, node
and pipe classes for an application node and one render client. Please
note that \textsf{Config::finishFrame} does block until the current
frame has been released locally and until the frame \textsf{current -
  latency} has been released by all nodes. The window and channel
synchronization are similar and omitted for simplicity.

It is absolutely vital for the execution that \textsf{Node::start\-Fra\-me}
and \textsf{Node::release\-Fra\-me} are called, respectively. The default
implementation of the node task methods does take care of that.

Since \textsf{eqPly} multi-buffers all dynamic data, it completely
removes frame synchronization by:

\begin{itemize}
\item releasing the local synchronization early in \textsf{Node::frameStart}
\item not waiting for the node to start the frame by not calling
  \textsf{Node::waitFrame\-Star\-ted} in \textsf{Pipe::frameStart}
\item not waiting for the pipe synchronization in
  \textsf{Node::frameDrawFinish} by not calling \textsf{Pipe::waitFrameLocal}
\end{itemize}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

@@

\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

@@

@@

A node represents a machine in the cluster, and is one process. It has a
name, a number of connection descriptions and at least one pipe. The
name of the node can be used for debugging, it has no influence on the
execution of Equalizer. For a description of connection attributes
please refer to \sref{sGlobal}.

{\footnotesize\begin{lstlisting}
node
{
    name     string
    connection // 0-n times, possible connections to this node
    {
        type          TCPIP | SDP
        TCPIP_port    unsigned
        hostname      string
        command       string      // render client launch command
        command_quote 'character' // launch command argument quote char
        timeout       unsigned    // timeout in milliseconds for launch
    }
    pipe {...} // 1-n times
}
\end{lstlisting}}

@@

>>>>

\usepackage{tabularx,longtable,graphicx,a4,listings,wrapfig,subfigure,textcomp}
\usepackage[absolute]{textpos}

@@

\date{
  \vspace{15cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.16 for Equalizer 0.6-RC1\\[\medskipamount]
  \today
}

@@

\thispagestyle{empty}
\maketitle
\begin{textblock}{}(0,5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock}

@@

\subsection*{Front and Back Page}

@@

The images on the back page show the following scalable rendering modes:
2D (sort-first, SFR), DB (sort-last), stereo, pixel and DPlex
(time-multiplex, AFR).

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes\\
    1.3.8   & Aug 25, 2008 & Update \sref{sEventHandling} with OSWindow changes\\
    1.3.9   & Aug 29, 2008 & Update \sref{sEventHandling}: event handling cleanup\\
    1.3.10  & Oct 7, 2008  & Updated to reflect 2D pixel compounds\\
    1.3.11  & Oct 13, 2008 & Updated to reflect eq::net::Object cleanup\\
    1.3.12  & Oct 16, 2008 & Added \sref{sView}\\
    1.3.13  & Oct 17, 2008 & Minor update in \sref{sMessagePump}\\
    1.3.14  & Oct 21, 2008 & Improved \sref{sConnectionDescription}\\
    1.3.15  & Oct 27, 2008 & Added \sref{sDPlex}\\
    1.3.16  & Nov 10, 2008 & Updated \sref{sThreadModel}, added back page\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains a current context for each
thread. If structured correctly, the application rarely has to take care
of thread synchonization or protection of shared data.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/model.pdf}
  {\caption{\small\label{fModel}Simplified execution model}}
\end{wrapfigure}
The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

Rendering threads on a single node are by default synchronized. When a
frame is finished, all local rendering threads are done
drawing. Therefore the application can safely modify the data between
the end of a frame and the beginning of a new frame. Furthermore, only
one instance of the scene data has to be maintained within a process
since all rendering threads are guaranteed to draw the same frame.

This per-node frame synchronization does not inhibit latency across
rendering nodes. Furthermore, advanced rendering software which
multi-buffers the dynamic parts of the database can disable the per-node
frame synchronization, as explained in \sref{sThreadModel}. Some scene
graphs implement multi-buffered data, and can profit from relaxing the
frame synchronization.

@@

By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of exising rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sThreadModel}.

@@

Equalizer applications use multiple, asynchronous execution threads. The
default execution model is focused on making the porting of existing
applications as easy as possible, as described in \sref{sExecution}. The
default, per-node thread synchronization provided by Equalizer can be
relaxed by advanced applications to gain better performance through
higher asynchronicity.

@@

\subsubsection{\label{sThreadModel}Thread Synchronization Models}
Equalizer supports three threading models, which can be set
programmatically by the application or through the configuration file
format. Applications typically hard-code their threading model. The file
format is commonly used to change the threading model for benchmarking
and experimentation.

The following thread synchronization models are implemented:
\begin{description}
\item[ASYNC:] No synchronization happens between render threads, except
  for synchonizing the finish of frame \textsf{current-latency}. The
  \textsf{eqPly} and \textsf{eVolve} Equalizer examples use this
  threading model, by setting it in \textsf{Node::configInit}. This
  synchonization model provides the best performance and should be used
  by all applications which multi-buffer all non-static, frame-specific
  data.
\item[DRAW\_SYNC:] In addition to async thread model, all local render
  threads are synchronized, so that the draw operations happen
  synchronously with the node main loop. This model allows to use the
  same database for rendering, and safe modifications of this database
  are possible from the node thread, since the pipe threads do not
  execute any rendering tasks between frames. This is the default
  threading model. This threading model should be used by applications
  which keep one copy of the scene graph per node.
\item[LOCAL\_SYNC:] In addition to the async threas model, all local
  frame operations, including readback, assemble and swap buffer are
  synchronized with the node main loop. This threading model should be
  used by applications which need to access non-buffered, frame-specific
  data after rendering, e.g., during \textsf{Channel::frameAssemble}.
\end{description}

\begin{figure}[ht!]\center
  \includegraphics[width=.9\textwidth]{images/threadModels.pdf}
  {\caption{\small\label{fThreadModels}Async, draw\_sync and local\_sync
    thread sunchronization models}}
\end{figure}

\fig{fThreadModels} illustrates the synchronization and task execution
for the thread synchronization models. Please note that the thread
synchronization synchronizes all pipe render threads on a
\textbf{single} node with the node's main thread. The per-node frame
synchronization does not break the asynchronous execution across
nodes.

\subsubsection{Thread Synchronization in Detail}

@@

The local synchronization consists of \textsf{releaseFrameLocal} to
unlock the local frame, and of \textsf{waitFrameLocal} to wait for the
unlock. For the default synchronization model \textsf{sync\_draw},
Equalizer uses the task method \textsf{frameDrawFinish} which is called
on each resource after the last \textsf{Channel::frameDraw} invocation
for this frame. Consequently, \textsf{Pipe::frameDraw\-Fi\-nish} calls
\textsf{Pipe::\-release\-Frame\-Lo\-cal} to signal that it is done
drawing the current frame, and \textsf{Node::frameDrawFinish} calls
\textsf{Pipe::waitFrameLocal} for each of its pipes to block the node
thread until the current frame has been drawn.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
\fig{fFrameSync} illustrates the local frame synchronization.

@@

\fig{fMainLoop} outlines the synchronization for the application, node
and pipe classes for an application node and one render client when
using the default \textsf{draw\_sync} thread model. Please note that
\textsf{Config::finishFrame} does block until the current frame has been
released locally and until the frame \textsf{current - latency} has been
released by all nodes. The window and channel synchronization are
similar and omitted for simplicity.

It is absolutely vital for the execution that \textsf{Node::start\-Fra\-me}
and \textsf{Node::release\-Fra\-me} are called, respectively. The default
implementations of the node task methods do take care of that.

@@

The main idea behind direct send is to parallelize the costly
recomposition for database (sort-last) decomposition. With each
additional source channel, the amount of pixel data to be composited
grows linearly. When using the simple approach of compositing all frames
on the destination channel, this channel quickly becomes the bottleneck
in the system. Direct send distributes this workload evenly across all
source channels, and thereby keeps the compositing work per channel
constant.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
  \vspace{-4ex}
\end{wrapfigure}
\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

@@

\begin{tabularx}{\textwidth}{|l|X|}
  \hline
  \textbf{Name} & EQ\_NODE\_IATTR\_THREAD\_MODEL\\
  \textbf{Value} & ASYNC \textbar \ DRAW\_SYNC \textbar \ LOCAL\_SYNC\\
  \textbf{Default} & DRAW\_SYNC\\
  \textbf{Details} & The threading model for node synchronization. See 
  \sref{sThreadModel} for details.\\
  \hline
\end{tabularx}\\\vfill

@@

A node represents a machine in the cluster, and is one process. It has a
name, a number of connection descriptions and at least one pipe. The
name of the node can be used for debugging, it has no influence on the
execution of Equalizer. For a description of node and connection
attributes please refer to \sref{sGlobal}.

{\footnotesize\begin{lstlisting}
node
{
    name     string
    connection // 0-n times, possible connections to this node
    {
        type          TCPIP | SDP
        TCPIP_port    unsigned
        hostname      string
        command       string      // render client launch command
        command_quote 'character' // launch command argument quote char
        timeout       unsigned    // timeout in milliseconds for launch
    }
    attributes
    {
        thread_model ASYNC | DRAW_SYNC | LOCAL_SYNC
    }
    pipe {...} // 1-n times
}
\end{lstlisting}}

@@

\cleardoublepage
\thispagestyle{empty}
\begin{textblock}{}(0,5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaserBack.pdf}
\end{textblock}
